---
title: "Temperature_Responses_CC"
author: "Dries Lorré"
date: "2025-09-26"
output: html_document
---
# Load required libraries
```{r, echo = FALSE}
library(ggplot2)
library(dplyr)
```

# Read data
```{r}
Temp_resp <- read.csv("Temp_sal_depth_ECOSIM_format_MS_NaN_removed.csv")
SST_m <- read.csv("SST_m.csv")
```
First, add column Temperature to Temp_resp to enable look-up of monthly temperatures: minimum value = 0, maximum value = 40 (This was done manually in Excel)
Next, link observed monthly temperature to environmental response (foraging compensation) of each species.
```{r}
# Loop through each species to interpolate responses
species_names <- colnames(Temp_resp)[c(4:30)]  # Get all species column names, excluding 'Temperature'

# Add columns for each species to SST_m (initializing with NA for now)
species_response_time_series <- SST_m

# Add empty columns for each species
for (species in species_names) {
  species_response_time_series[[species]] <- NA_real_
}

for (species in species_names) {
  # Use approx() for linear interpolation of species response
  interpolated_responses <- approx(Temp_resp$SST_month, Temp_resp[[species]], SST_m$SST_month)$y
  
  # Add the interpolated responses to the SST_m dataframe
  species_response_time_series[[species]] <- interpolated_responses
}

# Now species_response_time_series contains the SST_m time series with the interpolated species responses.