---
title: "Catch Time series for EwE (SBNS model)"
author: "Dries Lorré"
date: "2024-07-18"
output: html_document
---

Get packages
```{r setup, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(readxl)
```

Reading all data
```{r}
# ICES catch datasets
C621 <- read_xlsx("2006-2021_Catches.xlsx")
C5010 <- read_xlsx("ICES_HIST_CATCH_1950-2010.xlsx")
IVc_fisheries_catch <- read_xlsx("ICES_HIST_CATCH_1950-2010.xlsx")

# Species list
Species621 <- read_xlsx("SpeciesList_2006-2021.xlsx")
EwE_FG <- read.csv("totalcatchspecies.csv")
```

```{r}
area_IVc_km2 <- 63633.85614 #km2

IVc_fisheries_catch <- IVc_fisheries_catch[,-3]
IVc_fisheries_catch <- IVc_fisheries_catch %>%
  mutate_at(vars(3:ncol(IVc_fisheries_catch)), ~replace(.,. == "." | . == "-", 0)) %>%
  mutate_at(vars(3:ncol(IVc_fisheries_catch)), ~replace(.,. == "<0.5", 0.5)) %>%
  mutate_at(vars(3:ncol(IVc_fisheries_catch)), as.numeric)
  

#assign species from IVc_fisheries_catch not occuring in total landings species list to one that does
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Cuttlefish,bobtail squids nei")] <- "Cuttlefish, bobtail squids nei"
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Crangon shrimps nei")] <- "Common shrimp"
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Red mullet")] <- "Surmullets(=Red mullets) nei"
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Aesop shrimp")] <- "Pandalus shrimps nei"
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Clupeoids nei")] <- "Herrings, sardines nei"
IVc_fisheries_catch$Species[which(IVc_fisheries_catch$Species=="Scallops nei")] <- "Queen scallop"

#write.csv(rownames(total_catch),"data/totalcatchspecies.csv")  # list with all the caught species names according to the IVc_fisheries_catch df

# assign species to EwE FGs
EwE_FG$Species <- as.character(EwE_FG$Species)
EwE_FG$EwE_FG <- as.character(EwE_FG$EwE_FG)


# check if all species are in FGs or excluded
check <- NULL
IVc_fisheries_catch$EwE_FG <- NA
for (i in c(1:length(IVc_fisheries_catch$Species))){
  match <- which(IVc_fisheries_catch$Species[i]==EwE_FG$Species)
  if (length(match)==0){
    check <- rbind(check,i)
    next
  }
  IVc_fisheries_catch$EwE_FG[i] <- as.character(EwE_FG$EwE_FG[match])
}

#######################################################################
# All these species are not in FG or excluded -> change manual OR add to excel list OR exclude all (now as FG NA)
IVc_fisheries_catch$Species[check]

IVc_fisheries_catch$EwE_FG[which(IVc_fisheries_catch$Species=="Norway pout")] <- "Other gadoids"
IVc_fisheries_catch$EwE_FG[which(IVc_fisheries_catch$EwE_FG=="rays")] <- "Rays"
IVc_fisheries_catch$EwE_FG[94] <- "ND"  # Greenland halibut
IVc_fisheries_catch$EwE_FG[which(IVc_fisheries_catch$Species=="Sturgeons nei")] <- "ND"

# deze lijn zet ze voorlopig in FG ND dus worden eruit gefilterd
IVc_fisheries_catch$EwE_FG[which(is.na(IVc_fisheries_catch$EwE_FG))] <- "ND"

#######################################################################
# FISHERIES LANDINGS 1991 using landings from 2003 -----------------------------------------
#species total landings
subset_year <- IVc_fisheries_catch

total_landings <- array(NA,dim= c(length(unique(subset_year$EwE_FG)),61))
row.names(total_landings) <- c(unique(subset_year$EwE_FG))
colnames(total_landings) <- c(1950:2010)

for(sp in unique(subset_year$EwE_FG)){
  subset_species <- subset(subset_year,subset_year$EwE_FG==sp)
  total_landings[sp,] <- colSums(subset_species[,c(3:63)])
}
total_landings <- as.data.frame(total_landings)
```
Do not forget to correct for netherlands catches in some of the time series (manual)?
see excel!


Time series for 2006 till 2021.
First join the species list to the species codes and subseqently the EwE functional groups.
```{r}
C621[,5:ncol(C621)] <- lapply(C621[,5:ncol(C621)], as.numeric)

# C621 <- C621 %>%
#   rename(FAO_code = Species)

#combine the species names to the species codes
C621_species <- left_join(C621, Species621, by = c("Species" = "FAO_code"))

#now select the same species as in the historical data.
unique(C621_species$English)

C621_speciesEwE <- left_join(C621_species, EwE_FG, by = c("English" = "Species"))
```

List all the species that have not yet been assigned to an EwE functional group.
```{r}
# Historical dataset (1950-2010)
Species_5010 <- read.csv("soorten_1950-2010.csv", stringsAsFactors = FALSE)

# Recent dataset (2006-2021)
na_ewE_FG <- C621_speciesEwE[is.na(C621_speciesEwE$EwE_FG), ]
Species_0621 <- as.data.frame(unique(na_ewE_FG$English))
Species_0621 <- Species_0621 %>%
  rename(Species = `unique(na_ewE_FG$English)`)
# write.csv(Species_0621, "Species0621_NO_EwE_FG.csv")

Species_full <- full_join(Species_0621, Species_5010, by = "Species")
write.csv(Species_full, "Species_full.csv", row.names = FALSE)
```


```{r}
Species_EwE_FG <- read.csv("Species_EwE_FG.csv")
C621 <- read_xlsx("2006-2021_Catches.xlsx")
C5010 <- read_xlsx("ICES_HIST_CATCH_1950-2010.xlsx")

C5010 <- C5010[,-3]
C5010 <- C5010 %>%
  mutate_at(vars(3:ncol(C5010)), ~replace(.,. == "." | . == "-", 0)) %>%
  mutate_at(vars(3:ncol(C5010)), ~replace(.,. == "<0.5", 0.5)) %>%
  mutate_at(vars(3:ncol(C5010)), as.numeric)

C621[,5:ncol(C621)] <- lapply(C621[,5:ncol(C621)], as.numeric)

# Assign EwE functional groups to catch datasets
#combine the species names to the species codes
C621_species <- left_join(C621, Species621, by = c("Species" = "FAO_code"))
C621_speciesEwE <- left_join(C621_species, Species_EwE_FG, by = c("English" = "Species"))
C5010_SpeciesEwE <- left_join(C5010, Species_EwE_FG, by = "Species")

#check what species are not assigned an EwE FG:
missing_species <- setdiff(C621_speciesEwE$English, Species_EwE_FG$Species) #No missing species
missing_species <- setdiff(C5010_SpeciesEwE$Species, Species_EwE_FG$Species) #No missing species

# Get the total landings

landings_5010 <- C5010_SpeciesEwE %>%
  group_by(EwE_FG) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
landings_5010 <- as.data.frame(t(landings_5010))

Landings_621 <- C621_speciesEwE %>%
  group_by(EwE_FG) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
Landings_621 <- as.data.frame(t(Landings_621))


write.csv(landings_5010, "Landings_1950-2010.csv", row.names = FALSE)
write.csv(Landings_621, "landings_2006-2021.csv", row.names = FALSE)


```

add the corrections for other flatfish: correct for netherlands catch!

```{r}
C5010_SpeciesEwE_OF <- C5010_SpeciesEwE %>%
  filter(EwE_FG == "Other flatfish") %>%
  select(-c('1950':'1990'))

C5010_SpeciesEwE_OG <- C5010_SpeciesEwE %>%
  filter(EwE_FG == "Other gadoids") %>%
  select(-c('1950':'1990'))

C5010_SpeciesEwE_PF <- C5010_SpeciesEwE %>%
  filter(EwE_FG == "Pelagic fish") %>%
  select(-c('1950':'1990'))

C5010_SpeciesEwE_DF <- C5010_SpeciesEwE %>%
  filter(EwE_FG == "Demersal fish") %>%
  select(-c('1950':'1990'))

C5010_SpeciesEwE_SC <- C5010_SpeciesEwE %>%
  filter(EwE_FG == "Squid & cuttlefish")

write.csv(C5010_SpeciesEwE_SC, "Landings_Squidcuttlefish_91-10.csv", row.names = FALSE)

C621_speciesEwE_SC <- C621_speciesEwE %>%
  filter(EwE_FG == "Squid & cuttlefish")

write.csv(C621_speciesEwE_SC, "Landings_Squidcuttlefish_06-21.csv", row.names = FALSE)

```













































